---
title: "Probability of causation in individual workers with disease: Lung cancer due to occupational exposure to asbestos"
author: 
  - "Javier Mancilla-Galindo¹, Susan Peters¹, Jenny Deng¹, Henk van der Molen², Hans Kromhout¹, Lützen Portengen¹, Roel Vermeulen¹, Dick Heederik¹˒³"
  - "1.	Institute for Risk Assessment Sciences, Utrecht University, Utrecht, the Netherlands"
  - "2.	Department of Public and Occupational Health, Amsterdam UMC, Amsterdam, the Netherlands"
  - "3.	National Expertise Centre for Substance-related Occupational Diseases (Lexces), Utrecht, the Netherlands"
abstract: |
  <b>Background:</b> A lifetime cumulative exposure to asbestos of 25 fibre-years (4% risk increase per fibre-year) is used as the threshold at which lung cancer risk is twice as large in the Helsinki criteria, but studies have found steeper exposure-response at lower exposures. We aimed to determine occupational exposure thresholds where lung cancer is more likely than not caused by asbestos for workers’ compensation.  
  <b>Methods:</b> We estimated the relative risk (RR) of asbestos-related lung cancer by two approaches: spline meta-regression (spline-MR) of pooled RR estimates from systematic reviews (22 studies, 98633 participants), and meta-analysis of individual participant data of the SYNERGY pooled case-control study (14 studies, 37866 participants) with mixed-effects logistic regression adjusted for age, sex, smoking, and study. PoC was calculated as (RR-1)/RR for point estimates and upper 95% prediction intervals (PI), where PoC of 0.5 corresponds to doubling of lung cancer risk.
  <b>Results:</b> There was a 6.8% (PI: 0%–17.5%) lung cancer risk increase per fibre-year with minimal between-study heterogeneity (I² = 0.07%) in SYNERGY; 9.3% (PI: 2.7%–16.2%) in studies with population-based controls. Exposure-response was lower in spline-MR with substantial heterogeneity (I² = 92.7%). PoC equaled 0.5 at 81.7 (point estimate) and 26.1 fibre-years (upper PI) in spline-MR, compared to 10.6 and 4.3, respectively, in SYNERGY.
  <b>Conclusions:</b> The Netherlands adopted a 4.31 fibre-years threshold for asbestos-related lung cancer compensation, incorporating uncertainties in favor of claimants. SYNERGY estimates provided more consistent exposure-response estimates than meta-analysis of published aggregate data, supporting lower compensation thresholds than traditional Helsinki criteria.
keywords: ["Asbestos", "Lung cancer", "Probability of causation", "Causality", "Workers' compensation"]
execute: 
  echo: false
  warning: false
toc: false
format: 
  docx:
    reference-doc: template.docx
    link-citations: true
  pdf: 
    toc: true
    documentclass: scrartcl
  html:
    embed-resources: true
    toc: true
bibliography: PoC-Asbestos.bib
csl: bmj.csl
editor: visual
---

```{r}
if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,
  flextable,
  htmltools
  )
```

{{< pagebreak >}}

### Keywords

-   [**Asbestos**](https://www.ncbi.nlm.nih.gov/mesh/68001194)

-   [**Lung cancer**](https://www.ncbi.nlm.nih.gov/mesh/68008175)

-   **Probability of Causation**

-   [**Causality**](https://www.ncbi.nlm.nih.gov/mesh/68015984)

-   [**Workers' Compensation**](https://www.ncbi.nlm.nih.gov/mesh/68014942)

------------------------------------------------------------------------

### About this manuscript

Target journal: [*European Respiratory Journal (ERJ)*](https://publications.ersnet.org/content/erj)

Word count: 3119 (limit 3000)

Tables/Illustrations: 2 figures, 2 tables (max. 8 combined)

References: 36 (limit 40)

{{< pagebreak >}}

# Introduction

Determining whether an association between occupational exposure and disease is causal requires the integration of evidence from epidemiological studies, experimental animal data, and mechanistic research, including toxicological findings.[@Samet2020] Epidemiological information from population studies plays an important role because human evidence is critical for classification of causal agents as proven human carcinogens. Lung cancer has multiple known causes, with cigarette smoking being the leading risk factor.[@Leiter2023] There are various occupational exposures that increase the risk of lung cancer, including asbestos, respirable crystalline silica, diesel engine exhaust, arsenic, nickel, and chromium which have been classified as respiratory carcinogens.[@Leiter2023; @IARCChemicalAgentsOccupations2012] Different causal factors result in similar types of lung cancer, limiting the ability to conclude what caused lung cancer in an individual, for instance on the basis of tumour histology. Thus, assessment of individual causality is complicated for multi-causal diseases.

The only approach is to calculate the likelihood that in a lung cancer case, a certain determinant (chemical exposure) caused the disease. A commonly applied concept is the probability of causation (PoC).[@Dawid2017; @Tian2000] When PoC in an individual is greater than 0.5 (or 50%), the disease is “more likely than not” caused by the exposure of interest, as this corresponds mathematically to a doubling in the risk of the disease, meaning that the excess risk equals baseline risk (i.e., risk in non-exposed). This method has been applied in tort law and cancer compensation due to radiation,[@ProZES2020; @Wakeford2006] coal tar pitch volatiles,[@Armstrong1996] smoking,[@Siemiatycki2014] among others. While early critiques focused on its omission of accelerated outcome occurrence and non-identifiability from observational data,[@GreenlandPoC1999] later work showed that the PoC is identifiable under a causal model, monotonicity (no outcome prevention), and exogeneity (no confounding).[@Tian2000; @Li2024]

The PoC can be estimated from the exposure-response relation as observed in epidemiological studies or obtained from meta-analyses.[@Siemiatycki2014] However, it is sensitive to estimation errors arising from limitations in study design, conduct and analysis, such as exposure or disease measurement error and misclassification, confounding, causal model misspecification, and estimate imprecision, which may act to the detriment of the claimant. Therefore, it has been argued that uncertainties should be considered to the benefit of the claimant when applying the PoC.[@Wakeford2006; @Armstrong1996] Sources of error in asbestos exposure assessment include limited detection capability when sampling and analysis of fibres was not yet fully developed, difficulty in applying consistent fibre-to-mass or fibre-type conversion factors in case of exposure studies conducted using different measurement techniques,[@Franken2025] and incomplete or imprecise job histories, together generally resulting in underestimated or attenuated exposure-response, which has been estimated to be as large as a factor of 1.5 to 2.[@Lenters2011; @ECHA2021]

In the Netherlands, a compensation scheme for occupational diseases is in place since 2022, which adopts the *presumably plausible principle (‘voorshands aannemelijk principe’)*, operationalised by considering uncertainties in individual causality in favor of the worker, for a one-time compensation (all or none).[@Coenen2024] Lung cancer due to asbestos was one of the first occupational diseases included in the scheme in January 2022.

Asbestos was banned in the Netherlands in 1993; however, it was projected that 6,800 to 17,500 cases of asbestos-related lung cancer would occur from 2011 to 2030.[@vanderbij2016] The 2014 Helsinki report uses a lifetime cumulative exposure to asbestos of 25 fibre-years as the value of a doubling in the risk of lung cancer, which assumes an increase in risk of 4% per fibre-year of exposure.[@HelsinkiAsbestos2015] However, this estimate has been criticized for its lack of methodological transparency.[@Baur2017] For instance, the value of 4% has not been based on a formal meta-analysis of the available evidence. The Helsinki approach also assumes a linear exposure response relation which is inconsistent with more recent studies that indicate that the exposure-response may be steeper at lower exposure and levels off at higher exposure. Moreover, the slope of the exposure-response relation is significantly affected by the quality of the exposure assessment component of a study.[@Lenters2011; @vanderbij2013; @ECHA2021] In this study, we compare two approaches for estimating the level of occupational asbestos exposure associated with a doubling in the risk of lung cancer (PoC = 0.5): A meta-regression of pooled estimates obtained through a recently published systematic literature review and used in occupational standard setting in Europe, and data from a large meta-analysis of a multi-country case-control study with individual participant data which included workers who have been more recently exposed to asbestos and has a high-quality occupational exposure assessment component.

{{< pagebreak >}}

# Methods

## Study design and setting

### European Chemicals Agency (ECHA) study

```{r}
N_ECHA <- c(11000, 1056, 3513, 3072, 5770, 6358, 820, 1121, 1862, 13460, 740, 6931,
            2898, 1973, 1074, 17800, 4024, 3211, 2024, 577, 1539, 3296, 4514)

ECHA_absestos_exposure <- read.table("../../results/output_tables/ECHA_absestos_exposure.txt")
```

The ECHA evaluation made use of the systematic review by Lenters, et al.[@Lenters2011] (19 studies) which was updated by van der Bij, et al.[@vanderbij2013] and subsequently by ECHA for the so called Annex 1 connected to the Opinion of the Risk Assessment Committee on the occupational exposure limits for asbestos[@ECHA2021] (22 studies, N = `r sum(N_ECHA) %>% as.character` participants). Most studies (**Supplementary Table 1**) are occupational cohorts estimating standardised mortality ratios and only two estimated smoking-adjusted relative risks. The asbestos exposure period ranges from 1890 to 2009 (16 studies with exposure starting before World War II). Reported risk estimates correspond to cumulative asbestos exposures up to `r ECHA_absestos_exposure$Max` fibre-years (median `r ECHA_absestos_exposure$Median`, IQR: `r ECHA_absestos_exposure$Q1` – `r ECHA_absestos_exposure$Q3`). Exposure quality was rated high in 7 of the studies used by ECHA and low in 15 due to poor documentation, low exposure contrast, use of external fibre conversion factors, incomplete job history, or incomplete exposure coverage.[@Lenters2011]

### SYNERGY study

```{r}
#| include: false
summary_table <- read.table("../../results/output_tables/summary_continuous_exposure.txt")

median_cum0_overall <- summary_table$asbestos_cum0_median[3] %>% round(2)
q1_cum0_overall <- summary_table$asbestos_cum0_q1[3] %>% round(2)
q3_cum0_overall <- summary_table$asbestos_cum0_q3[3] %>% round(2)
min_cum0_overall <- summary_table$asbestos_cum0_min[3] 
max_cum0_overall <- summary_table$asbestos_cum0_max[3] %>% round(2)
```

The SYNERGY project is a pooled case-control study on the risk of different occupational carcinogens for the development of lung cancer. SYNERGY collected data from 14 case-control studies conducted across Europe and in Canada (1985–2010), comprising a total of 17,705 lung cancer cases and 21,813 matched controls (**Supplementary Table 2**). Lung cancer cases with histological or cytological confirmation were included in the individual case-control studies with an overall response rate of 83% (range: 62–98%). Controls are mostly recruited from the general population (79% vs 21% hospitals), with a response rate of 70% (range: 41–100%). A nested case-control study within MORGEN, and the LUCA and ICARE studies used density sampling, while the remaining studies employed cumulative incidence sampling with frequency matching on age and sex. Odds ratios can thus be interpreted as rate and risk ratios, respectively, which converge[@kerr2023] because lung cancer is relatively rare and asbestos exposure prevalence stays stable over the short recruitment windows relative to the long latency of asbestos.

Cumulative occupational exposure to asbestos was estimated using the quantitative job exposure matrix (JEM) *SYN-JEM*,[@Peters2016SYNJEM] which uses the region, duration, and period of exposure within an ISCO-68 job title to estimate exposure to asbestos in fibre-years (ff/ml-years). The model used to create the JEM incorporated the effect of any country-specific asbestos bans and it was applied in a large epidemiological study of the relationship between asbestos and lung cancer.[@Olsson2017] Cumulative asbestos exposure in SYNERGY ranges from `r min_cum0_overall` to `r max_cum0_overall` fibre-years (median `r median_cum0_overall`, IQR: `r q1_cum0_overall` – `r q3_cum0_overall`).

The SYNERGY study was approved by the IARC Ethics Committee in addition to the individual study ethics approval.

## Statistical Analysis

### Exposure-response estimation

The exposure-response relation in the ECHA report was estimated with random-effects meta-regression of the study-level logRR and average exposure level per exposure category reported. Four models were examined (linear and natural splines, allowing for differences or no differences in background lung cancer risk). Predictions were computed over a range of 0–200 fibre-years and plotted. The RR was calculated from meta-regression fixed-effects estimates. The spline meta-regression model with fixed intercept was used for further analysis due to reasons explained elsewhere,[@vanderbij2013] for the estimation of $RR(x)$ using the natural spline (NS) basis functions (Equation 1).

$$
RR(x) = \exp\left(\beta_1 \cdot NS_1(x) + \beta_2 \cdot NS_2(x)\right)
\quad (1)
$$

A meta-analysis of SYNERGY with individual participant data was used to estimate the RR of lung cancer due to lifetime cumulative asbestos based on the posited causal model described in the **Supplementary Methods** and represented in a directed acyclic graph (**Supplementary Figure 1**, <https://dagitty.net/mjsrVdmjg>). Meta-estimates were obtained with a generalized linear mixed model (logit link) with study-specific random intercepts and random slopes for exposure to account for potential between-study heterogeneity based on the different SYNERGY study results. The model was adjusted for sex, study, age group (\<45, 45–49, 50–54, 55–59, 60–64, 65–69, 70–74, and \>74 years), smoking (cigarette pack-years \[log (cigarette pack-years +1)\] and time-since-quitting smoking cigarettes (current smokers; stopping smoking 2–7, 8–15, 16–25, ≥26 years before interview/diagnosis; and never-smokers). $RR(x)$ was obtained from the model population-level fixed effect $\gamma$ (Equation 2).

$$
RR(x) = \exp\left(\gamma \cdot x\right)
\quad (2)
$$

### Heterogeneity and prediction intervals

Between-study heterogeneity was quantified using the I² statistic, representing the proportion of total variation attributable to genuine differences between studies rather than sampling error. In ECHA, I² reflected heterogeneity in reported logRR. In SYNERGY, heterogeneity in the exposure-response slope was expressed as $I^2 = \tau^2_{\text{slope}} / (\tau^2_{\text{slope}} + \pi^2 / 3) \times 100\%$, where $\tau^2_{\text{slope}}$ is the between-study variance in slopes and $\pi^2 / 3$ approximates within-study variance.[@Nakagawa2017] Higher I² values indicate greater inconsistency in estimated effects across studies beyond what would be expected by chance.

For both the ECHA meta-regression and SYNERGY analysis, 95% prediction intervals (PI) were calculated using Higgins, et al.[@HigginsMetanalysis2009] method, which combines the squared standard error of the target effect with its estimated between-study variance. In ECHA, the target was the pooled logRR and in SYNERGY, the asbestos exposure-response slope coefficient. PI represent the range in which the true effect estimate from a new study would be expected to fall, accounting for both sampling uncertainty and between-study heterogeneity.

### Probability of Causation

PoC for an individual case is equal to the difference between risk under exposure ($\text{Risk}_{\text{exp}}$) minus baseline risk ($\text{Risk}_{\text{unexp}}$), divided by $\text{Risk}_{\text{exp}}$ (Equation 3), which can be re-expressed as relative risks (RR) (Equation 4).[@Tian2000]

<!-- PoC expressed in absolute risks -->

$$
PoC = \frac{Risk_{exp} - Risk_{unexp}}{Risk_{exp}}
\quad (3)
$$

<!-- PoC expressed as relative risk -->

$$
PoC = \frac{RR - 1}{RR}
\quad (4)
$$

For quantitative exposures, PoC can be estimated using the RR at any given exposure value ($x$) from the exposure-response relation (Equation 5).[@Armstrong1996; @Siemiatycki2014]

<!-- PoC for quantitative exposure -->

$$
{PoC}(x) \;=\; \frac{RR(x)\;-\;1}{RR(x)} \quad (5)
$$

We estimated the PoC across the range of cumulative asbestos exposure in SYNERGY. Equation 5 was solved for $PoC(x) = 0.5$ for both exposure-response relations to obtain exposure thresholds for compensation. The frequency of lung cancer cases with a calculated PoC ≥ 0.5 is reported as the total count of compensated cases and the rate of compensated cases per 10,000 lung cancer cases using the SYNERGY population as reference.

Besides PoC calculated with the $RR(x)$ point estimates, upper limits of the 95% PI were also used in the calculation of PoC to account for uncertainties in the exposure-response relations (*presumably plausible* threshold).

### Sensitivity Analysis

The exposure-response relations in SYNERGY were re-estimated stratifying by source of controls (hospital-based and population-based), since hospital-based controls are known to have greater baseline disease risk than population-based controls, resulting in potential underestimation of the effect. Exposure-response and PoC thresholds are presented as previously described.

All analyses and visualizations were performed in R v.4.4.0 through RStudio v.2024.12.1. The analysis code and documentation are available through <https://github.com/UtrechtUniversity/PoC-Asbestos>.

{{< pagebreak >}}

# Results

```{r}
#| include: false
contingency_table <- read.table("../../results/output_tables/contingency_table.txt")

cases <- sum(contingency_table$Cases)
controls <- sum(contingency_table$Controls)

exposed_cases <- contingency_table$Cases[1]
exposed_controls <- contingency_table$Controls[1]

nonexposed_cases <- contingency_table$Cases[2]
nonexposed_controls <- contingency_table$Controls[2]

total_population <- cases + controls
total_exposed <- exposed_cases + exposed_controls

or <- (exposed_cases/exposed_controls)/(nonexposed_cases/nonexposed_controls)
af <- (or-1)/or
```

A summary of the characteristics of the SYNERGY study population (N = `r total_population`) is provided in **Table 1**. Ever exposure to asbestos was reported in `r round((total_exposed/total_population*100), 1)`% (n = `r total_exposed`) of the participating individuals. The percentage of exposure to asbestos in lung cancer cases and controls was `r round((exposed_cases/cases*100), 1)`% (n = `r exposed_cases`/`r cases`) and `r round((exposed_controls/controls*100), 1)`% (n = `r exposed_controls`/`r controls`), respectively.

**Table 1**. Descriptive characteristics of SYNERGY study participants.

```{r}
includeHTML("tables/Table1.html")
```

> Abbreviations: fibre-years (ff/ml-years), maximum (Max), minimum (Min), 25th percentile (Q1), 75th percentile (Q3), standard deviation (SD).

{{< pagebreak >}}

```{r}
#| include: false

median_cum0_cases <- summary_table$asbestos_cum0_median[2] %>% round(2)
q1_cum0_cases <- summary_table$asbestos_cum0_q1[2] %>% round(2)
q3_cum0_cases <- summary_table$asbestos_cum0_q3[2] %>% round(2)
min_cum0_cases <- summary_table$asbestos_cum0_min[2] 
max_cum0_cases <- summary_table$asbestos_cum0_max[2] %>% round(2)

median_cum0_controls <- summary_table$asbestos_cum0_median[1] %>% round(2)
q1_cum0_controls <- summary_table$asbestos_cum0_q1[1] %>% round(2)
q3_cum0_controls <- summary_table$asbestos_cum0_q3[1] %>% round(2)
min_cum0_controls <- summary_table$asbestos_cum0_min[1] 
max_cum0_controls <- summary_table$asbestos_cum0_max[1] %>% round(2)

median_dur0_cases <- summary_table$asbestos_dur0_median[2] 
q1_dur0_cases <- summary_table$asbestos_dur0_q1[2] 
q3_dur0_cases <- summary_table$asbestos_dur0_q3[2]
min_dur0_cases <- summary_table$asbestos_dur0_min[2] 
max_dur0_cases <- summary_table$asbestos_dur0_max[2] 

median_dur0_controls <- summary_table$asbestos_dur0_median[1] 
q1_dur0_controls <- summary_table$asbestos_dur0_q1[1] 
q3_dur0_controls <- summary_table$asbestos_dur0_q3[1] 
min_dur0_controls <- summary_table$asbestos_dur0_min[1] 
max_dur0_controls <- summary_table$asbestos_dur0_max[1] 
```

The median lifetime cumulative exposure to asbestos in ever-exposed was `r median_cum0_cases` (IQR: `r q1_cum0_cases` – `r q3_cum0_cases`, range: `r min_cum0_cases` – `r max_cum0_cases`) fibre-years in cases and `r median_cum0_controls` (IQR: `r q1_cum0_controls` – `r q3_cum0_controls`, range: `r min_cum0_controls` – `r max_cum0_controls`) fibre-years in controls. Total duration of exposure in the cases and controls was `r median_dur0_cases` (IQR: `r q1_dur0_cases` – `r q3_dur0_cases`, range: `r min_dur0_cases` – `r max_dur0_cases`) years and `r median_dur0_controls` (IQR: `r q1_dur0_controls` – `r q3_dur0_controls`, range: `r min_dur0_controls` – `r max_dur0_controls`) years, respectively.

### Exposure-response and PoC in ECHA

Exposure-response relations of RR of lung cancer with increasing level of asbestos exposure, highlighting the range of exposure values (0 to 65 fibre-years) observed in SYNERGY, are shown in **Supplementary Figure 2**. Using the spline model with no intercept (I² = 92.7%), lung cancer RR is not doubled below 65 fibre-years with the point estimate, while the upper limit of the 95% PI does (**Figure 1**). PoC of 0.5 occurs at 81.7 and 26.1 fibre-years, respectively (**Figure 2A**). Zero and three lung cancer cases per 10,000 would be entitled to compensation under the point estimate and presumably plausible thresholds based on ECHA (**Table 2**).

```{r}
counts_spline <- read.table("../../results/output_tables/PoC_spline_model_cum0_count_plus50.txt")

spline_upper_bound_rate <- counts_spline %>% pull(Presumably_plausible_per_10k)

spline_upper_bound_min.cum.exp <- counts_spline %>% pull(Min_exp_conservative)
```

![**Figure 1**. Exposure response relations of lung cancer risk increase at low cumulative asbestos exposure. Solid lines represent the average meta-estimates; shaded areas, the upper 95% prediction intervals, and the dotted red line, the doubling in increase of risk threshold (RR = 2). Vertical tick marks represent individual participant exposure values in SYNERGY (blue, top) and average exposure values for the reported pooled estimates per exposure catgory in ECHA (green, bottom).](images/Figure_1.png){fig-align="center"}

{{< pagebreak >}}

**Table 2**. Number of compensated cases per 10,000 lung cancer cases according to various scenarios applied to the lung cancer cases (n = 16901) within the SYNERGY population, for the probability of causation (PoC) due to lifetime cumulative exposure to asbestos (fibre-years). It was estimated at what cumulative exposure the PoC is equal or larger than 50%.

```{r}
includeHTML("tables/Table2.html") 
```

{{< pagebreak >}}

### Exposure-response and PoC in SYNERGY

The exposure-response relation in SYNERGY (I² = 0.07%) estimates a 6.8% lung cancer risk increase per fibre-year, with an upper PI of 17.5% (**Figure 1**). This corresponds to a PoC of 0.5 (**Figure 2B**) at 10.6 fibre-years for the point estimate and 4.3 under the presumably plausible threshold (**Table 2**), for 115 and 845 compensated cases per 10,000, respectively. In sensitivity analysis, a steeper exposure-response (**Supplementary Figure 3**) was observed for studies with population-based controls (9.3% risk increase per fibre-year) compared to studies with hospital controls (1.9% risk increase per fibre-year), with large shifts in the PoC = 0.5 thresholds (**Supplementary Figure 4**) for the point estimates, but limited impact in the presumably plausible threshold for population controls compared to the average SYNERGY population effect (**Table 2**).

{{< pagebreak >}}

![**Figure 2**. Probability of causation of lung cancer across the range of lifetime cumulative exposure to asbestos (fibre-years) values in SYNERGY, based on **A)** exposure-response relation in the ECHA study and **B)** the SYNERGY study. Solid lines represent the average meta-estimates; shaded areas, the upper 95% prediction interval (Upper PI), and the dotted blue line, the PoC = 0.5 threshold. SYNERGY estimates were adjusted for age, sex, smoking (pack-years), and time since quitting smoking.](images/Figure_2.png){fig-align="center"}

{{< pagebreak >}}

# Discussion

In this study, we applied two approaches for the calculation of the probability of causation of lung cancer due to exposure to asbestos, to determine exposure thresholds for the financial compensation of lung cancer cases occupationally exposed to asbestos. We used a widely applied method for the calculation of PoC[@Armstrong1996; @Wakeford1998; @Siemiatycki2014] from the exposure-response relation of lung cancer risk due to lifetime cumulative exposure to asbestos of (1) a systematic review with spline meta-regression [@Lenters2011; @vanderbij2013; @ECHA2021] and (2) a large pooled case-control study of lung cancer and occupational carcinogens (SYNERGY).[@Olsson2017] Our scenario analyses explore and incorporate uncertainties to the benefit of the claimant through the use of a presumably plausible principle.[@Coenen2024] These findings show that the lifetime cumulative exposure threshold of 4.3 fibre-years based on SYNERGY to compensate workers occupationally exposed to asbestos is defensible, as it incorporates uncertainties to the benefit of the claimant.

Siemiatycki, et al. recommend a two stages for the calculation of the PoC, where the critical amount of exposure at which PoC = 0.5 (RR = 2) is estimated from the exposure-response curve from a published meta-analysis of studies in multiple populations, or individual epidemiological studies when the first is not available or feasible.[@Siemiatycki2014] As a second step, the fraction of cases exceeding the critical value of exposure can be estimated, requiring that the exposure metric is available for both steps. We chose lifetime cumulative exposure to asbestos determined from SYN-JEM as it is strongly associated with lung cancer, and meets the condition for the two-stepped approach.

One important limitation of the ECHA spline meta-regression is that it underestimates the exposure-response curve due to low quality exposure assessment in some studies, a widely recognized issue due to exposure misclassification and measurement error.[@Grandjean2004] Other potential sources of bias in ECHA are confounding due unmeasured confounders and possibly selection bias due to healthy worker effect in occupational cohorts, the latter most often resulting in effect underestimation.[@Consonni2025] The large between-study heterogeneity (I² = 92.7%) in ECHA suggests that uncertainties in the exposure-response relation are due to true between-study differences in design, conduct, and bias.

Our second approach used the exposure-response estimated through meta-analysis of individual SYNERGY participant data. Cumulative exposure to asbestos in SYNERGY is based on SYN-JEM,[@Peters2016SYNJEM] which has the advantage of assigning values with a homogeneous method across studies. However, it introduces Berkson-type error by assigning average exposure values,[@ArmstrongMeasurementError1998] resulting in imprecise (wider confidence intervals) estimations of the exposure-response curve.[@Rhomberg2011] Contrary to ECHA, there is little heterogeneity of effects between studies in SYNERGY (I² = 92.7%). While there is for exposure misclassification, for instance, if the job title does not fully represent the average exposure for the specific workers' tasks, such misclassification is non-differential with respect to outcome due to the independent dedicated development of SYN-JEM. While most measurements used to develop SYN-JEM used phase contrast microscopy (PCM) with a lower limit of detection (LOD) of \~2,000 fibres/m³, 5% of measurements were with electron microscopy (EM) with LOD 100 to 200 fibres/m³.[@Franken2025] In contrast, none of the studies in ECHA used EM, whereas many applied conversion factors due to the changes over time in analytic techniques, thus highlighting the higher exposure quality in SYNERGY.[@Lenters2011]

In the Dutch context, uncertainties are used in the favor of the worker to receive a one-time compensation for occupational diseases. Our approach incorporates substantial knowledge of the exposure-response relationship to define exposure thresholds of PoC = 50% at the upper bound of statistical uncertainty (95% PI). However, extensive documentation may not be available for other occupational diseases and exposures. Alternative ways of incorporating uncertainties may be used in such cases. For instance, lowering of the PoC threshold had been used in the United Kingdom to initiate partial payments at PoC of 20% and full payment above 50%,[@Wakeford2006] and Italy, where flexible use of PoC thresholds of 25% and 50% has been proposed according to the specific judicial scenario.[@AIRMpoc2017] Both cases use the upper bound of the 95%CI, which is different from the approach used in the US where greater weight is given to statistical uncertainties by using the 99%CI.[@EEOICPA2020] Lack of knowledge of the exposure-response curve in the population of interest can be a source of uncertainty in other cases, such as the German compensation scheme for cancer due to radiation, where dose-response predictions derived from radiation exposure in atomic bomb survivors in Hiroshima and Nagasaki are corrected to be applicable in the population of Germany.[@ProZES2020] Nonetheless, other sources of uncertainty remain largely unexplored, such as incorporation of potential residual confounding and selection into the estimations.

Strengths of this study include the large sample size and diverse representativeness of European and Canadian populations. The PoC can be calculated from epidemiological studies and is easy to communicate to occupational health workers, patients, and stakeholders. Nonetheless, incorporation of uncertainties into its calculation to benefit the individual worker affected by an occupational disease requires substantial knowledge of the sources of uncertainty. More recently, decision-theory approaches have been proposed as alternatives to counterfactual-based estimates such as the PoC.[@Sarvet2025] However, part of this debate remains largely theoretical, as both approaches would yield equivalent results when either detects harm.[@Sarvet2025a] Potential future directions for legal applications and workers' compensation could involve other health outcomes accounting for accelerated occurrence and different dimensions of affection, such as years of working life lost,[@YWLL2021] disability-adjusted life years,[@Haagsma2015] or loss of income.[@HernandezIncomeLoss2021] These approaches, however, also rely on average population effects to make inferences about the individual.

{{< pagebreak >}}

# Conclusions

This study shows that individual participant data meta-analysis provides more reliable exposure-response estimates of lung cancer due to asbestos than traditional meta-regression of published studies for workers' compensation. Steeper exposure-response relations in SYNERGY with minimal between-study heterogeneity translate into lower compensation thresholds under the presumably plausible principle. Therefore, an exposure threshold of 4.3 fibre-years has been selected for the one-time compensation of workers occupationally exposed to asbestos in the Netherlands who develop lung cancer. The newer estimates based on SYNERGY should supersede those based on the Helsinki criteria.

## Funding

This work was supported by the Dutch Ministry of Social Affairs and Employment, grant number 25719.

## Competing interests

None reported.

## Data and code availability

Data are available upon reasonable request (inquiries can be directed to Dr. Susan Peters. Alternatively, <https://synergy.iarc.who.int/contact/>).

The source code for this manuscript, statistical analyses, and project documentation is available through the online repository: <https://github.com/UtrechtUniversity/PoC-Asbestos>.

{{< pagebreak >}}

# References
