---
title: "Probability of Causation: ECHA Systematic Review with Meta-regression"
subtitle: "Lung cancer due to occupational asbestos exposure"
author:
  - name: Javier Mancilla-Galindo
    affiliations:
      - ref: iras
  - name: Susan Peters
    affiliations:
      - ref: iras
  - name: Jenny Deng
    affiliations:
      - ref: iras
  - name: Henk van der Molen
    affiliations:
      - ref: amc
  - name: Hans Kromhout
    affiliations:
      - ref: iras
  - name: Lützen Portengen
    affiliations:
      - ref: iras
  - name: Roel Vermeulen
    affiliations:
      - ref: iras
  - name: Dick Heederik
    affiliations:
      - ref: iras
      - ref: lexces
affiliations:
  - id: iras
    name: Institute for Risk Assessment Sciences, Utrecht University
    city: Utrecht
    country: the Netherlands
  - id: amc
    name: Department of Public and Occupational Health, Amsterdam UMC
    city: Amsterdam
    country: the Netherlands
  - id: lexces
    name: National Expertise Centre for Substance-related Occupational Diseases (Lexces)
    city: Utrecht
    country: the Netherlands
date: today
execute: 
  echo: false
  warning: false
format:
  pdf: 
    toc: true
    colorlinks: true
    documentclass: scrartcl
bibliography: ../docs/manuscript/PoC-Asbestos.bib
csl: ../docs/manuscript/european-respiratory-journal.csl
editor: source
---

\pagebreak

```{r}
#| label: directories
#| include: false

# Create directories for sub-folders  
inputfolder <- "../data/raw"
psfolder <- "../data/processed"
tempfolder <- "../data/temp"
figfolder <- "../results/output_figures"
tabfolder <- "../results/output_tables"
manuscript_images <- "../docs/manuscript/images"
manuscript_tables <- "../docs/manuscript/tables"

dir.create(inputfolder, showWarnings = FALSE)
dir.create(psfolder, showWarnings = FALSE)
dir.create(tempfolder, showWarnings = FALSE)
dir.create(figfolder, showWarnings = FALSE)
dir.create(tabfolder, showWarnings = FALSE)
dir.create(manuscript_images, showWarnings = FALSE)
dir.create(manuscript_tables, showWarnings = FALSE)
```

```{r}
#| label: packages
#| include: false 

if (!require("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

pacman::p_load(
  tidyverse,        # Used for basic data handling and visualization.
  conflicted,       # Resolve conflicts between packages.
  table1,           # Create table of descriptive characteristics. 
  flextable,        # Used to render and save table 1. 
  lme4,             # Linear mixed-effects models.
  mixmeta,          # Derive exposure-response function based on meta-analysis.
  splines,          # Natural spline to model exposure-response function.
  gridExtra,        # Arrange multiple ggplot2 plots.
  docstring,        # Write and read documentation of functions. 
  gt,               # Used to print html tables.  
  report            # Used to cite packages used in this session.   
)

conflicts_prefer(dplyr::filter, dplyr::select, dplyr::lag)
```

```{r}
#| label: load data
#| include: false 

source("scripts/load_data.R")
```

\pagebreak


```{r}
exposed <- df %>% filter(ever_asbestos0 == 1) %>% count %>% pull
cases <- df %>% filter(status == 1) %>% count %>% pull

cases_male <- df %>% filter(status == 1 & sex == "Male") %>% count %>% pull
cases_female <- df %>% filter(status == 1 & sex == "Female") %>% count %>% pull
```

## Spline Model

This model incorporates the the exposure-response relation from nonlinear meta-regression models as determined by Lenters, et al.[@Lenters2011] and van der Bij, et al.[@vanderbij2013], and updated in the scientific opinion for the European Chemicals Agency (ECHA).[@ECHA2021] The underlying data and results of the meta-regression are contained within the list element `MOD`. Tailored-made functions are sourced into this analysis, namely:

-   `predict.nsplin`: Generates predictions using natural spline transformations based on the provided model object.
-   `getRR`: Calculate relative risk (RR) based on a specified (linear/spline) model that estimates the exposure-response relationship.
-   `getPoC`: Calculate probability of causation (PoC) based on the RR obtained from getRR function.


```{r}
#| echo: true

MOD <- readRDS(file.path(inputfolder,"MOD.rds"))
source("scripts/predict.nsplin.R")
source("scripts/getRR.R")
source("scripts/getPoC.R")
```


```{r}
#| eval: false

# Documentation of functions: 
docstring(predict.nsplin)
docstring(getRR)
docstring(getPoC)
```

There are 4 different models possible:

1.  linear model, assumes difference in background rate of outcome

2.  linear model, assumes no difference in background rate of outcome

3.  spline model, assumes difference in background rate of outcome

4.  spline model, assumes no difference in background rate of outcome

```{r}
#| include: false 
# pull out the 4 models and their data in lists
models   <- map(MOD, ~ .x$m[[1]])
datasets <- map(MOD, "data")

# loop by index
walk(seq_along(models), function(i) {
  mod_i <- models[[i]]
  data_i <- datasets[[i]]

  # console summary
  cat(sprintf("\n------- Model %d -------\n", i))
  print(summary(mod_i))

  # save data_i as txt
  write.table(data_i, 
              file = file.path(tempfolder, sprintf("data_model_%d.txt", i)),
              sep = "\t",
              row.names = FALSE,
              quote = FALSE)
})
```

```{r}
ECHA_absestos_exposure <- datasets[[4]] %>% 
  filter(exposure > 0) %>% 
  summarize(
    N = n(),
    Min = min(exposure),
    Q1 = quantile(exposure, 0.25),
    Median = median(exposure),
    Q3 = quantile(exposure, 0.75),
    Max = max(exposure)
  ) %>% round(1)

ECHA_absestos_exposure %>% 
  write.table(., paste0(tabfolder, "/ECHA_absestos_exposure.txt"), sep = "\t")
```


The range of reported exposure values greater than 0 for the pooled estimates in ECHA is `r ECHA_absestos_exposure$Min` - `r ECHA_absestos_exposure$Max` fiber-years, with a median of `r ECHA_absestos_exposure$Median` (IQR: `r ECHA_absestos_exposure$Q1` - `r ECHA_absestos_exposure$Q3`)

\pagebreak

```{r}
I2 <- summary(models[[4]])$i2stat
```

The I-square statistic for model 4 (spline, no difference in background rate) is **`r I2 %>% round(1)`%**. The following plot shows the relative risk estimates reported by study, grouped by quality of exposure assessment. The plot data margins were fixed to y = 5 and x = 100 to focus on lower exposure values, thus resulting in the removal of 52 rows out of 134. 

```{r}
datasets[[4]] %>% mutate(quality = factor(quality, levels = c("0","1"), labels = c("low", "high"))) %>% 
ggplot(., aes(x = exposure, y = exp(logRR), group = id, color = quality)) +
  geom_point(alpha = 0.3, size = 1.5) +
  geom_line() +
  
  labs(
    x = "Fiber-years (ff/ml-years)",
    y = "RR",
    color = "Quality",
    title = "Exposure-Response Curves by Study Quality"
  ) +
  xlim(0,100) +
  ylim(0,5) +
  scale_color_brewer(palette = "Set1") +  # Use a colorblind-friendly palette
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  ) + 
  facet_wrap(~ quality)
```

\pagebreak

### Exposure-response 

```{r}
#| include: false
source("scripts/plot_spline.R")
```

Shape of the exposure response under the 0-200 fiber-years range: 

```{r}
spline_plot
```

This figure shows that a doubling in the risk of lung cancer (RR = 2) is not expected to be seen with the values of exposure observed in the SYNERGY study (shaded area).


Using modelling strategy number 4, and the actual exposure data in SYNERGY, the PoC and prediction intervals as a measure of uncertainty from study heterogeneity can be obtained following Higgins, et al. method.[@HigginsMetanalysis2009]

\pagebreak

This is the plot of lung cancer risk ratio with the observed exposure values in this study: 

```{r}
#| warning: false
spline_RR
```

And the probability of causation, with the upper prediction interval: 

```{r}
#| warning: false
spline_PoC 
```

\pagebreak

Participants with outcome and a PoC greater than or equal to 50%:

```{r}
beta_coefficients_spline <- df_grid %>%
  summarise(
    Min_exp_50pct_PoC = round(asbestos_cum0[which(PoC >= 0.5)[1]], 2),
    Min_exp_conservative = round(asbestos_cum0[which(PoC_PI_upper >= 0.5)[1]], 2),
    .groups = 'drop'
  ) %>%
  # Add risk estimates at 1.5 fibre-years
  bind_cols(
    df_grid %>%
      filter(abs(asbestos_cum0 - 1.5) == min(abs(asbestos_cum0 - 1.5))) %>%
      summarise(
        Risk_per_ffyr_instantaneous = first(Risk_per_ffyr_instantaneous),
        Risk_per_ffyr_upper_instantaneous = first(Risk_per_ffyr_upper_instantaneous),
        Risk_per_ffyr_lower_instantaneous = first(Risk_per_ffyr_lower_instantaneous)
      )
  ) %>%
  mutate(
    Model = "Spline",
    Risk_per_ffyr_est = paste0(round(Risk_per_ffyr_instantaneous, 1), "%"),
    Risk_per_ffyr_PI = paste0(round(Risk_per_ffyr_lower_instantaneous, 1), "% ; ", round(Risk_per_ffyr_upper_instantaneous, 1), "%")
  ) %>%
  # Add case counts
  mutate(
    Point_per_10k = round(sum(df$status == 1 & df$asbestos_cum0 >= Min_exp_50pct_PoC, na.rm = TRUE) / cases * 10000),
    Presumably_plausible_per_10k = round(sum(df$status == 1 & df$asbestos_cum0 >= Min_exp_conservative, na.rm = TRUE) / cases * 10000)
  ) %>%
  select(Model, Risk_per_ffyr_est, Risk_per_ffyr_PI, Min_exp_50pct_PoC, Min_exp_conservative, Point_per_10k, Presumably_plausible_per_10k)

beta_coefficients_spline %>% 
  write.table(., paste0(tabfolder, "/PoC_spline_model_cum0_count_plus50.txt"), sep = "\t")
```

```{r}
beta_coefficients_spline <- beta_coefficients_spline %>% 
  gt() %>%
  cols_label(
    Model = "Model",
    Risk_per_ffyr_est = "Estimate",
    Risk_per_ffyr_PI = "95% Prediction Interval", 
    Min_exp_50pct_PoC = "Point Estimate",
    Min_exp_conservative = "Presumably Plausible",
    Point_per_10k = "Point Estimate", 
    Presumably_plausible_per_10k = "Presumably Plausible"
  ) %>%
  tab_spanner(
    label = "Risk Increase per Fibre-year",
    columns = c(Risk_per_ffyr_est, Risk_per_ffyr_PI)
  ) %>%
  tab_spanner(
    label = "Min Exposure for 50% PoC (fibre-years)",
    columns = c(Min_exp_50pct_PoC, Min_exp_conservative)
  ) %>%
  tab_spanner(
    label = "Cases per 10,000 above 50% PoC in SYNERGY",
    columns = c(Point_per_10k, Presumably_plausible_per_10k)
  ) %>%
  cols_align(align = "center") %>%
  tab_header(
    title = "Asbestos Exposure-Response in ECHA (Spline Model)"
  ) %>%
  tab_footnote(
    footnote = "Probability of Causation (PoC)",
    locations = cells_column_spanners(spanners = "Min Exposure for 50% PoC (fibre-years)")
  ) %>%
  tab_footnote(
    footnote = "Not applicable for non-linear spline model",
    locations = cells_column_spanners(spanners = "Risk Increase per Fibre-year")
  ) %>% 
  tab_footnote(
    footnote = "Instantaneous slope at 1.5 fibre-years (median exposure in SYNERGY cases)",
    locations = cells_column_spanners(spanners = "Risk Increase per Fibre-year")
  ) 

# Save table
beta_coefficients_spline %>%  
  gtsave(filename = "rtf/PoC_spline_model_cum0_count_plus50.rtf", path = tabfolder)

beta_coefficients_spline %>% 
  cols_width(
    everything() ~ pct(15) 
  )
```


\pagebreak


# References

::: {#refs}
:::

\pagebreak

```{r}
#| label: session
# remove clutter
session <- sessionInfo()
session$BLAS <- NULL
session$LAPACK <- NULL
session$loadedOnly <- NULL
# write log file
writeLines(
  capture.output(print(session, locale = FALSE)),
  paste0("sessions/",lubridate::today(), "_PoC_Asbestos_spline.txt")
)                                   
```

# Package References

For specific information on the operating system, R version, and R package versions used, please refer to the **R/session** folder in the GitHub repository.

```{r}
#| output: asis
report::cite_packages(session)
```

```{r}
#| include: false

# Run this chunk if you wish to clear your environment and unload packages.

pacman::p_unload(negate = TRUE)

rm(list = ls())
```
